Assessing primitives performance\\ on multi-stage execution

### 2018-05-10 paper.tex
minor fixes with Sophie

For this reason, some virtual machine implementors attempted to remove most or all optional primitives. \sd{sentence was too long: please reread my version:}Their goal was to use a JIT performing speculative optimizations of optional primitives implemented in the client language. Ideally we wanted to get   the performance of the same primitives written in the VM. Self \cite{ursPHD} and Strongtalk \cite{Strongtalk} were \sk{the?} to first to try this approach. Early versions of the Javascript engine V8 \cite{V8} also implemented most of the primitives, including for example Array operations in Javascript itself. However, without optional primitives implemented in the VM, several problems rise:
\begin{itemize}
	\item High performance relies on a JIT with speculative optimizations, which is hard to implement, maintain and evolve,
	\item The performance of the code not yet optimized is much worse than the peak performance code,
	\item Even mature JIT fail to optimize some narrow cases, where the performance is drastically slower.
\end{itemize}
Overall, this approach requires high engineering time to get a mature optimizing JIT with speculative optimizations and even then, it leads in practice to unreliable performance.

@@

VM plugins enable the addition of new features to the VM. There are 2 types of plugins: the built-ins and the externals.
\begin{itemize}
\item External plugins. Plugins that can be added or removed from the VM without the need to recompile the VM sources.
\item Built-in plugins. Plugins that cannot be installed without recompiling the VM sources. However, these plugins can reach some parts of the VM internals\footnote{as data structures, routines...} that remains unreachable from an external plugin. These operations are performed via an InterpreterProxy. Although, this solution results in a substential overhead if called often.
\end{itemize}

@@

\cba{We may use paragraph instead of itemize here to compact}
 In this section we compare the five different implementations according to the following criteria:
 \begin{itemize}
 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes).
	\item \emph{Implementation language familiar to developers:} Smalltalk application developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use \emph{All} if all communities are familiar with the language, \emph{VM} if all VM developers are and \emph{VM-JIT} if only a subset of VM developers are.
	\item \emph{Readability:} To evaluate the readability of the approach, we use one of the most used metric, the number of lines of code.
	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to execute if the primitive fails. This code is maintain by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation. However, as a VM implementor, we need to maintain what generates the C code of the VM, hence, in the case of Smart Syntax, we need to maintain a different method in each client language. We also count all the Slang implementations as well as the ones on Cog's RTL.
	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
	\item \emph{Optimization potential:} \cba{need to mention that this is for large strings}
	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through InterpreterProxy to do so, leading to an overhead.
	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT will always generate the same native code.
\end{itemize}

@@

\paragraph{GCC-Linux.}The GCC-Linux evaluation was performed on an Asus Zenbook with Ubuntu 16.04 LTS, a 2,2 Ghz processor Intel Core 5, 8 Gb 1600MHz DDR3 of RAM. The Linux VM was compiled using GCC version 5.4.0.

\paragraph{LLVM-Mac.}The LLVM-Mac evaluation was performed on a MacBook pro with Mac OS 10.11.6, a 2.9Ghz processor Intel Core 5, 8 Gb 1867MHz DDR3 of RAM. The Mac VM was compiled using Apple LLVM version 8.0.0 (clang-800.0.38).

@@

In each benchmark, we run a string comparison in a loop and measure the total execution time.
The number of iterations of the loop depends on the string size: the comparison is repeated between 1414 and 10000 times for a short string (5 characters or less), and between 100 and 316 times for a long string (100 characters or more). This addresses the warm-up of the JIT \cba{??? This needs to be explained differently. We get into jitted code after a small number of loop iteration (I think 20 currently) or second call. The JIT overhead is negligible here for 1 second because we could not see any time spent in the JIT in the profiler + ref to your paper, i.e. less than 0.5ms in the JIT} and allows the benchmarks to run at least 1 second in the slower implementations \cba{Ok, 1 second to avoid processor / OS noises, not just for fun}.
The loop overhead is measured and removed from the measurements.

@@

\paragraph{Optimization potential on large strings comparison.}
The plugins come with an overhead \sk{not an overhead when it comes to inlining. Just not optimized. need rephrasing} due to the cross-file inlining and the passage through an InterpreterProxy, like said in Section \ref{sec:implem_cmp}. This is one of the factor explaining the performance gap between these versions and the Slang and Slang+JIT ones.
When it comes to comparing large strings (100 characters and more), the execution time is mainly spent in the comparison loop (Figure \ref{fig:Asm}). However, the switch overhead is still noticeable in the PureSlang implementation when strings of 100 characters are compared: PureSlang is 15.8 times faster while Slang+JIT is 16.8 times faster. This is not the case anymore on 1000-characters strings.
This last observation only applies to the GCC results: the code generated by the LLVM compiler induces a longer execution time, as explained in the next paragraph. \sk{is it legal to ref that way?} \cba{it's ok, but not recommended since sometimes we move things around without checking other sections}
\paragraph{Control over generated code.}
One could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, the performances between the PureSlang and Slang+JIT implementations are really close for long strings (100 characters and longer), with respectively 24,42 and 24,48 relative speed-up factors for a 1000-characters string. Yet they greatly differ when LLVM is used, with respectively 13,44 and 26,40 relative speed-up factors.

>>>>

For this reason, some virtual machine implementors attempted to remove most or all optional primitives. Their goal was to use a JIT performing speculative optimizations to be able to implement the optional primitives in the client language without performance loss. Ideally we wanted to get   the performance of the same primitives written in the VM. Self \cite{ursPHD} and Strongtalk \cite{Strongtalk} were the first to try this approach. Early versions of the Javascript engine V8 \cite{V8} also implemented most of the primitives, including for example Array operations in Javascript itself. However, without optional primitives implemented in the VM, several problems rise:
\begin{itemize}
	\item High performance relies on a JIT with speculative optimizations, which is hard to implement, maintain and evolve,
	\item The performance of the code not yet optimized is much worse than the peak performance code,
	\item Even mature JIT fail to optimize some narrow cases, where the performance is drastically slower.
\end{itemize}
Overall, this approach requires high engineering time to get a mature optimizing JIT with speculative optimizations and even then, it leads in practice to unreliable performance.

@@

VM plugins enable the addition of new features to the VM. There are 2 types of plugins: the built-ins and the externals.
\begin{itemize}
\item \emph{External plugins:} Plugins that can be added or removed from the VM without the need to recompile the VM sources.
\item \emph{Internal plugins:} Plugins that cannot be installed without recompiling the VM sources. However, these plugins can reach some parts of the VM internals\footnote{as data structures, routines...} that remains unreachable from an external plugin. These operations are performed via an InterpreterProxy. Although, this solution results in a substential overhead if called often. \cba{Remove from paper InterpreterProxy}
\end{itemize}

@@

 In this section we compare the five different implementations according to the following criteria:
 \begin{itemize}
 	\item \emph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes).
	\item \emph{Implementation language familiar to developers:} Smalltalk application developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use \emph{All} if all communities are familiar with the language, \emph{VM} if all VM developers are and \emph{VM-JIT} if only a subset of VM developers are.
	\item \emph{Readability:} To evaluate the readability of the approach, we use one of the most used metric, the number of lines of code.
	\item \emph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to execute if the primitive fails. This code is maintain by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation. However, as a VM implementor, we need to maintain what generates the C code of the VM, hence, in the case of Smart Syntax, we need to maintain a different method in each client language. We also count all the Slang implementations as well as the ones on Cog's RTL.
	\item \emph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
	\item \emph{Optimization potential:} \cba{need to mention that this is for large strings}
	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through InterpreterProxy to do so, leading to an overhead. \cba{InterpreterProxy not defined before}
	\item \emph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT will always generate the same native code.
\end{itemize}

@@

\paragraph{GCC-Linux.}The GCC-Linux evaluation was performed on an Asus Zenbook with Ubuntu 16.04 LTS, a 2.2 Ghz processor Intel Core 5, 8 Gb 1600MHz DDR3 of RAM. The Linux VM was compiled using GCC version 5.4.0.

\paragraph{LLVM-Mac.}The LLVM-Mac evaluation was performed on a MacBook pro with Mac OS 10.11.6, a 2.9 Ghz processor Intel Core 5, 8 Gb 1867MHz DDR3 of RAM. The Mac VM was compiled using Apple LLVM version 8.0.0 (clang-800.0.38).

@@

In each benchmark, we run a string comparison in a loop and measure the total execution time.
The number of iterations of the loop depends on the string size: the comparison is repeated between 1414 and 10000 times for a short string (5 characters or less), and between 100 and 316 times for a long string (100 characters or more). This addresses the warm-up of the JIT \cba{??? This needs to be explained differently. We get into jitted code after a small number of loop iteration (I think 20 currently) or second call. The JIT overhead is negligible here for 1 second because we could not see any time spent in the JIT in the profiler + ref to your paper, i.e. less than 0.5ms in the JIT} \cite{CogProfiler} and allows the benchmarks to run at least 1 second in the slower implementations \cba{Ok, 1 second to avoid processor / OS noises, not just for fun}.
The loop overhead is measured and removed from the measurements.

@@

\paragraph{Optimization potential on large strings comparison.}
The plugins come with an overhead \sk{not an overhead when it comes to inlining. Just not optimized. need rephrasing} due to the cross-file inlining and the passage through an InterpreterProxy, like said in Section \ref{sec:implem_cmp}. This is one of the factor explaining the performance gap between these versions and the Slang and Slang+JIT ones.
When it comes to comparing large strings (100 characters and more), the execution time is mainly spent in the comparison loop (Figure \ref{fig:Asm}). However, the switch overhead is still noticeable in the PureSlang implementation when strings of 100 characters are compared: PureSlang is 15.8 times faster while Slang+JIT is 16.8 times faster. This is not the case anymore on 1000-characters strings.
This last observation only applies to the GCC results: the code generated by the LLVM compiler induces a longer execution time, as explained in the next paragraph.

One could notice a rather striking performance difference between the 2 compilers: when compiling with GCC, the performances between the PureSlang and Slang+JIT implementations are really close for long strings (100 characters and longer), with respectively 24,42 and 24,48 relative speed-up factors for a 1000-characters string. Yet they greatly differ when LLVM is used, with respectively 13,44 and 26,40 relative speed-up factors.
